Modelling M/EEG data with Bayesian nonparametric multilevel models

### 2025-02-09 brms_meeg.tex
including some more refs and using t2 splines for timegen data

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021a},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al., 2019};
\citeproc{ref-pernet2015}{Pernet et al., 2015};
\citeproc{ref-riutort-mayol_practical_2023}{Riutort-Mayol et al., 2023};
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025})\ldots{} See also
(\citeproc{ref-maris2011}{Maris, 2011})\ldots{} and
(\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}) (history of
cluster-based approaches and using a data split?)\ldots{} Cluster
failure (\citeproc{ref-eklund2016}{Eklund et al., 2016})\ldots{}

@@

Disentangling overlapping processes
(\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021})\ldots{}
Using Bayes factors (\citeproc{ref-teichmann2022}{Teichmann,
2022})\ldots{}

@@

@@

Following the approach used by Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see details and code in
\citeproc{ref-yeung2004}{Yeung et al., 2004}). As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), the signal was generated
from truncated Gaussian with an objective onset at 160 ms, a peak at 250
ms, and an offset at 342 ms. We simulated this signal for 250 timesteps
between 0 and 0.5s, akin to \& 500 Hz sampling rate. We simulated such
data for a group of 20 participants with 50 trials per participant and
condition.

\caption{Some ERPs in two conditions with 50 trials each, for a group of
20 participants.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/eeg-group-1.pdf}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# averaging across participants}
\NormalTok{group\_df }\OtherTok{\textless{}{-}}\NormalTok{ raw\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{pivot\_wider}\NormalTok{(}\AttributeTok{names\_from =}\NormalTok{ condition, }\AttributeTok{values\_from =}\NormalTok{ y, }\AttributeTok{values\_fn =}\NormalTok{ mean) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{y\_diff =}\NormalTok{ cond2 }\SpecialCharTok{{-}}\NormalTok{ cond1) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarise}\NormalTok{(}
        \AttributeTok{y\_mean =} \FunctionTok{mean}\NormalTok{(y\_diff),}
        \AttributeTok{y\_sem =} \FunctionTok{sd}\NormalTok{(y\_diff) }\SpecialCharTok{/} \FunctionTok{sqrt}\NormalTok{(}\FunctionTok{n}\NormalTok{() ),}
        \AttributeTok{.by =}\NormalTok{ x}
\NormalTok{        ) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{lower =}\NormalTok{ y\_mean }\SpecialCharTok{{-}}\NormalTok{ y\_sem, }\AttributeTok{upper =}\NormalTok{ y\_mean }\SpecialCharTok{+}\NormalTok{ y\_sem)}

\CommentTok{\# plotting the data}
\NormalTok{group\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x, }\AttributeTok{y =}\NormalTok{ y\_mean) ) }\SpecialCharTok{+}
    \FunctionTok{geom\_hline}\NormalTok{(}\AttributeTok{yintercept =} \DecValTok{0}\NormalTok{, }\AttributeTok{linetype =} \DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{geom\_vline}\NormalTok{(}\AttributeTok{xintercept =}\NormalTok{ true\_onset, }\AttributeTok{linetype =} \DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{geom\_ribbon}\NormalTok{(}
        \FunctionTok{aes}\NormalTok{(}\AttributeTok{ymin =}\NormalTok{ lower, }\AttributeTok{ymax =}\NormalTok{ upper, }\AttributeTok{colour =} \ConstantTok{NULL}\NormalTok{),}
        \AttributeTok{alpha =} \FloatTok{0.25}\NormalTok{, }\AttributeTok{show.legend =} \ConstantTok{FALSE}
\NormalTok{        ) }\SpecialCharTok{+}
    \FunctionTok{geom\_line}\NormalTok{(}\AttributeTok{show.legend =} \ConstantTok{FALSE}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (ms)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG difference (a.u.)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\caption{Group-level average difference between conditions (mean +/-
standard error of the mean). The `true' (population value of the) onset
is indicated by the vertical dashed line.}

{\centering \includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/erp-diff-1.pdf}

@@

Models were fitted using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019})\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# defining a contrast for condition}
\FunctionTok{contrasts}\NormalTok{(group\_df}\SpecialCharTok{$}\NormalTok{condition) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{)}

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
    \CommentTok{\# then we should try s(x, participant, bs = "fs")}
    \CommentTok{\# that is, random factor smooth interactions...}
\NormalTok{    y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(x, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ group\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{gp\_model }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# k refers to the number of basis functions for}
    \CommentTok{\# computing Hilbert{-}space approximate GPs}
\NormalTok{    y }\SpecialCharTok{\textasciitilde{}} \FunctionTok{gp}\NormalTok{(x, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \CommentTok{\# if k = NA (default), exact GPs are computed}
    \CommentTok{\# y \textasciitilde{} gp(x, by = condition),}
    \AttributeTok{data =}\NormalTok{ group\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.99}\NormalTok{),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gp.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\FunctionTok{plot\_post\_preds}\NormalTok{(}\AttributeTok{model =}\NormalTok{ gam, }\AttributeTok{data =}\NormalTok{ group\_df) }\SpecialCharTok{+}
    \FunctionTok{plot\_post\_preds}\NormalTok{(}\AttributeTok{model =}\NormalTok{ gp\_model, }\AttributeTok{data =}\NormalTok{ group\_df)}
\end{Highlighting}
\end{Shaded}

@@

We can also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional BF thresholds (Figure~\ref{fig-post-prob-ratio}).

@@

Below we are plotting the error function (where the error is computed as
\(|\hat{\theta}-\theta|\)) for both the onset and offset values of the
ERP difference, according to various \texttt{eps} and \texttt{threshold}
values. Remember that the signal is generated from a truncated Gaussian
defining an objective onset at 160 ms, a maximum at 250 ms, and an
offset at 342 ms. Figure~\ref{fig-onset-error} shows that the GP model
can \emph{exactly} recover the true onset and offset values, given some
reasonable choice of \texttt{eps} and \texttt{threshold} values.

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.06        15             160              342           0            0
2 0.06        16             160              342           0            0
3 0.06        17             160              342           0            0
4 0.06        18             160              342           0            0
5 0.06        19             160              342           0            0
6 0.06        20             160              342           0            0
\end{verbatim}

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.06         3             159              342           1            0
2 0.06         4             159              342           1            0
3 0.06         5             159              342           1            0
4 0.06         6             159              342           1            0
5 0.06         7             159              342           1            0
6 0.06         8             159              342           1            0
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# averaging across participants}
\NormalTok{summary\_df }\OtherTok{\textless{}{-}}\NormalTok{ raw\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarise}\NormalTok{(}
        \AttributeTok{eeg =} \FunctionTok{mean}\NormalTok{(y),}
        \AttributeTok{eeg\_sd =} \FunctionTok{sd}\NormalTok{(y),}
        \AttributeTok{.by =} \FunctionTok{c}\NormalTok{(participant, condition, x)}
\NormalTok{        )}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(x, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gam, }\AttributeTok{effect =} \StringTok{"x:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (ms)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\caption{Hierarchical GAM posterior predictions.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/meta-gam-preds-1.pdf}

@@

\caption{\label{fig-sim-decoding}Exemplary (simulated) group-level
average timecourse of binary decoding accuracy (ROC AUC).}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-sim-decoding-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# loading the reticulate package}
\FunctionTok{library}\NormalTok{(reticulate)}

\CommentTok{\# importing the numpy and mne python modules}
\NormalTok{np }\OtherTok{\textless{}{-}} \FunctionTok{import}\NormalTok{(}\StringTok{"numpy"}\NormalTok{)}
\NormalTok{mne }\OtherTok{\textless{}{-}} \FunctionTok{import}\NormalTok{(}\StringTok{"mne"}\NormalTok{)}
\NormalTok{sklearn }\OtherTok{\textless{}{-}} \FunctionTok{import}\NormalTok{(}\StringTok{"sklearn"}\NormalTok{)}

\CommentTok{\# defining the function in R (it will be executed in Python)}
\NormalTok{mne\_decoding }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{ (X, labels, }\AttributeTok{ncores =} \DecValTok{1}\NormalTok{) \{}

>>>>

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al., 2019};
\citeproc{ref-pernet2015}{C. R. Pernet et al., 2015};
\citeproc{ref-riutort-mayol_practical_2023}{Riutort-Mayol et al., 2023};
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025})\ldots{} See also
(\citeproc{ref-maris2011}{Maris, 2011})\ldots{} and
(\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}) (history of
cluster-based approaches and using a data split?)\ldots{} Cluster
failure (\citeproc{ref-eklund2016}{Eklund et al., 2016})\ldots{}

@@

Disentangling overlapping processes
(\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021})\ldots{}
Using Bayes factors (\citeproc{ref-teichmann2022}{Teichmann,
2022})\ldots{} Weighting single trials (\citeproc{ref-pernet2022}{C.
Pernet, 2022})\ldots{}

See
\url{https://www.let.rug.nl/nerbonne/teach/rema-stats-meth-seminar/presentations/Wieling-2014-GAMs.pdf}
and \url{https://www.let.rug.nl/wieling/Statistics/GAM-EEG/lab/},
\url{https://www.let.rug.nl/wieling/Statistics/GAM-EEG/GAM-EEG.pdf}
(\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015})\ldots{}

Recent example of GLM for EEG (\citeproc{ref-fischer2013}{Fischer \&
Ullsperger, 2013}; \citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al.,
2025})\ldots{} ``Similar approaches have been successfully applied to
EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and
Cavanagh, 2011) data and allow the simultaneous investigation of
multiple independent variables while preserving the high temporal
resolution of the EEG.''\ldots{} See also (\citeproc{ref-hauk2006}{Hauk
et al., 2006}; \citeproc{ref-rousselet2008}{Rousselet et al.,
2008})\ldots{} Example of two-stage regression analysis
(individual-level then group-level, 10.1101/2024.04.10.588861,
\citeproc{ref-dunagan2024}{Dunagan et al., 2024})\ldots{}

@@

See also
\url{https://jtimonen.github.io/lgpr-usage/articles/math.html}\ldots{}

@@

Following the approach used by Sassenhagen \& Draschkow
(\citeproc{ref-sassenhagen2019}{2019}) and Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), we simulated EEG data
stemming from two conditions, one with noise only, and the other with
noise + signal. As in previous studies, the noise was generated by
superimposing 50 sinusoids at different frequencies, following an
EEG-like spectrum (see details and code in
\citeproc{ref-yeung2004}{Yeung et al., 2004}). As in Rousselet
(\citeproc{ref-rousselet_using_2025}{2025}), the signal was generated
from truncated Gaussian with an objective onset at 160 ms, a peak at 250
ms, and an offset at 342 ms. We simulated this signal for 250 timesteps
between 0 and 0.5s, akin to a 500 Hz sampling rate. We simulated such
data for a group of 20 participants with 50 trials per participant and
condition (Figure~\ref{fig-eeg}).

\caption{\label{fig-eeg}Some ERPs in two conditions with 50 trials each,
for a group of 20 participants.}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-eeg-1.pdf}

@@

We computed the average of the ERP difference
(Figure~\ref{fig-erp})\ldots{}

\caption{\label{fig-erp}Group-level average difference between
conditions (mean +/- standard error of the mean). The `true' (population
value of the) onset is indicated by the vertical dashed line.}

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-erp-1.pdf}

@@

We then fitted a GAM and GP regression model using the \texttt{brms}
package (\citeproc{ref-brms2017}{Bürkner, 2017},
\citeproc{ref-brms2018}{2018}; \citeproc{ref-nalborczyk2019}{Nalborczyk
et al., 2019})\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# averaging across participants}
\NormalTok{ppt\_df }\OtherTok{\textless{}{-}}\NormalTok{ raw\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(participant, condition, time) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{eeg =} \FunctionTok{mean}\NormalTok{(eeg) ) }\SpecialCharTok{\%\textgreater{}\%}
    \CommentTok{\# mutate(eeg = as.numeric(scale(x = eeg, scale = FALSE) ) ) \%\textgreater{}\%}
    \FunctionTok{ungroup}\NormalTok{()}

\CommentTok{\# defining a contrast for condition}
\FunctionTok{contrasts}\NormalTok{(ppt\_df}\SpecialCharTok{$}\NormalTok{condition) }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.5}\NormalTok{, }\FloatTok{0.5}\NormalTok{)}

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# computing eeg\_diff and averaging across participants}
\CommentTok{\# group\_df \textless{}{-} raw\_df \%\textgreater{}\%}
\CommentTok{\#     group\_by(participant, time) \%\textgreater{}\%}
\CommentTok{\#     pivot\_wider(names\_from = condition, values\_from = eeg, values\_fn = mean) \%\textgreater{}\%}
\CommentTok{\#     mutate(eeg\_diff = cond2 {-} cond1) \%\textgreater{}\%}
\CommentTok{\#     summarise(eeg\_diff = mean(eeg\_diff) ) \%\textgreater{}\%}
\CommentTok{\#     ungroup()}

\CommentTok{\# see https://betanalpha.github.io/assets/case\_studies/gp\_part3/part3.html}
\CommentTok{\# and https://betanalpha.github.io/assets/case\_studies/gaussian\_processes.html}
\NormalTok{gp\_priors }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"normal(0, 1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"Intercept"}\NormalTok{),}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"normal(0, 1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{),}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"exponential(1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"sigma"}\NormalTok{),}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"exponential(1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"sdgp"}\NormalTok{)}
\NormalTok{    )}

\NormalTok{gp\_model }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# k refers to the number of basis functions for}
    \CommentTok{\# computing Hilbert{-}space approximate GPs}
    \CommentTok{\# if k = NA (default), exact GPs are computed}
    \CommentTok{\# eeg \textasciitilde{} gp(time, by = condition),}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{gp}\NormalTok{(time, }\AttributeTok{by =}\NormalTok{ condition, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{cov =} \StringTok{"exp\_quad"}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \CommentTok{\# prior = gp\_priors,}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.99}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{15}\NormalTok{),}
    \AttributeTok{backend =} \StringTok{"cmdstanr"}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gp.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\FunctionTok{plot\_post\_preds}\NormalTok{(}\AttributeTok{model =}\NormalTok{ gam) }\SpecialCharTok{+} \FunctionTok{plot\_post\_preds}\NormalTok{(}\AttributeTok{model =}\NormalTok{ gp\_model)}
\end{Highlighting}
\end{Shaded}

@@

We can also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional BF thresholds (Figure~\ref{fig-post-prob-ratio}).
Also check this paper
(\url{https://link.springer.com/article/10.3758/s13428-021-01613-6?fromPaywallRec=false})?

@@

Below we are plotting the error function (where the error is computed as
\(\text{error}:=|\hat{\theta}-\theta|\)) for both the onset and offset
values of the ERP difference, according to various \texttt{eps} and
\texttt{threshold} values. Remember that the signal is generated from a
truncated Gaussian defining an objective onset at 160 ms, a maximum at
250 ms, and an offset at 342 ms. Figure~\ref{fig-onset-error} shows that
the GP model can \emph{exactly} recover the true onset and offset
values, given some reasonable choice of \texttt{eps} and
\texttt{threshold} values.

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.03         6            0.16            0.351           0        0.009
2 0.03         7            0.16            0.351           0        0.009
3 0.03         8            0.16            0.351           0        0.009
4 0.03         9            0.16            0.351           0        0.009
5 0.03        10            0.16            0.351           0        0.009
6 0.03        11            0.16            0.351           0        0.009
\end{verbatim}

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.07         8           0.166            0.342       0.006            0
2 0.07         9           0.166            0.342       0.006            0
3 0.07        10           0.166            0.342       0.006            0
4 0.07        11           0.166            0.342       0.006            0
5 0.07        12           0.166            0.342       0.006            0
6 0.07        13           0.166            0.342       0.006            0
\end{verbatim}

\begin{verbatim}
[1] eps              threshold        estimated_onset  estimated_peak  
[5] estimated_offset error_onset      error_peak       error_offset    
<0 rows> (or 0-length row.names)
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# averaging across participants}
\NormalTok{summary\_df }\OtherTok{\textless{}{-}}\NormalTok{ raw\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarise}\NormalTok{(}
        \AttributeTok{eeg\_mean =} \FunctionTok{mean}\NormalTok{(eeg),}
        \AttributeTok{eeg\_sd =} \FunctionTok{sd}\NormalTok{(eeg),}
        \AttributeTok{.by =} \FunctionTok{c}\NormalTok{(participant, condition, time)}
\NormalTok{        )}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting the GP}
\NormalTok{meta\_gp }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{gp}\NormalTok{(time, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.99}\NormalTok{),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gp.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\NormalTok{meta\_gam\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gam, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}
\NormalTok{meta\_gp\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gp, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}

\CommentTok{\# combining the two plots}
\NormalTok{meta\_gam\_preds }\SpecialCharTok{+}\NormalTok{ meta\_gp\_preds}
\end{Highlighting}
\end{Shaded}

@@

\caption{Posterior predictions for the hierarchical GAM and GP.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/meta-gp-preds-1.pdf}

@@

\caption{\label{fig-mne-decoding}Exemplary average timecourse of binary
decoding accuracy (ROC AUC) for each participant. Group-level average
decoding accuracy is depicted as a grey background density in each
panel.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-mne-decoding-1.pdf}
